### 2016-10-04 Paper_v7.tex
DB semi final

\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
\vspace{0.05in}
{\par \centering Russell Gould$^{\ddag}$ and David M. Bassen, Anirikh Chakrabarti, Jeffrey D. Varner and Jonathan Butcher$^{\ddag *}$}
\vspace{0.05in}
{\par \centering Robert Frederick Smith School of Chemical and Biomolecular Engineering and $^{\ddag}$Nancy E. and Peter C. Meinig School of Biomedical Engineering, Cornell University, Ithaca NY 14853}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~ Modeling of TGF-$\beta$ induced EMT}
\vspace{0.1in}
% {\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology (60,000 Character Limit)}}
%\vspace{0.1in}
{\par \centering \textbf{Character count:}~\emph{56,659}}
\vspace{0.1in}
{\par \centering $^{*}$Corresponding author:}
\vspace{0.1in}
{\par \centering Jonathan T. Butcher,}
{\par \centering Associate Professor, Nancy E. and Peter C. Meinig School of Biomedical Engineering,}
{\par \centering 304 Weill Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jtb47@cornell.edu}
{\par \centering Phone: (607) 255 - 3575}
{\par \centering Fax: (607) 255 - 7330}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
\section*{Abstract}
Epithelial to mesenchymal transition (EMT) is an essential differentiation program during tissue morphogenesis and remodeling.
EMT is induced by soluble transforming growth factor $\beta$ (TGF-$\beta$) family members, and restricted by vascular endothelial growth factor family members.
While many downstream molecular regulators of EMT have been identified, these have been largely evaluated individually without considering potential crosstalk.
In this study, we created an ensemble of dynamic mathematical models describing TGF-$\beta$ induced EMT to better understand the operational hierarchy of this complex molecular program.
These models incorporate mass action kinetics within an ordinary differential equation (ODE) framework to describe the transcriptional and post-translational regulatory events driving EMT.
Model parameters were estimated from multiple data sets using multiobjective optimization, in combination with cross-validation.
TGF-$\beta$ exposure drove the model population toward a mesenchymal phenotype, while an epithelial phenotype was maintained following vascular endothelial growth factor A (VEGF-A) exposure.
Simulations predicted that the transcription factors phosphorylated SP1 and NFAT were master regulators promoting or inhibiting EMT, respectively.
Surprisingly, simulations also predicted that a cellular population could exhibit phenotypic heterogeneity (characterized by a significant fraction of the population with both high epithelial and mesenchymal marker expression)
if treated simultaneously with TGF-$\beta$ and VEGF-A.  We tested this prediction experimentally in both MCF10A and DLD1 cells and found that upwards of 45\% of the cellular population acquired this hybrid state in the presence of both TGF-$\beta$ and VEGF-A.
We experimentally validated the predicted NFAT/Sp1 signaling axis for each phenotype response. Lastly, we found that cells in the hybrid state had significantly different functional behavior when compared to VEGF-A or TGF-$\beta$ treatment alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and potentially reveal a novel signaling axis which regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

We identified model subpopulations that exhibited different behaviors following exposure to TGF$-\beta$1/2 (Fig. \ref{fg:F3}B).
Analysis of the molecular signatures of these subpopulations suggested the abundance, localization and state of the Sp1, AP-1 and NFATc transcription factors controlled population heterogeneity.
The majority of models ($>$80\%) responded to treatment, moving away from the untreated population (Fig \ref{fg:F3}A-C, gray).
These models showed the classically expected behavior, a switch from an epithelial to mesenchymal phenotype following TGF$-\beta$1/2 exposure.
Some models resembled untreated cells; they had evaluated nuclear localized phosphorylated Sp1, relative to non-induced cells,
which decreased E-cadherin expression through Slug-mediated inhibition, which in turn increased Vimentin expression through TGF$-\beta$3 autocrine signaling and the liberation of $\beta$-catenin. However, the most biologically interesting behavior was exhibited by cells achieving a hybrid phenotype, most notable in a dual treatment condition (\ref{fg:F3}C, black arrow), but also present in a small percentage of untreated cells (Fig. \ref{fg:F3}A).
Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in simultaneously increased Vimentin and E-cadherin expression (Fig. \ref{fg:F4}A).

@@

\section*{Discussion}
In this study, we developed a family of mechanistic models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of VEGF-A.
The signaling architecture encoded in the model, which contained 74 molecular species interconnected by 169 interactions,
described the expression of 23 genes in response to growth factor stimulation.
This simulation incorporates an unprecedented level of detail compared to previous models, but as a consequence created a large number of unknown model parameters.
Because these parameters could not be estimated uniquely apriori, we estimated an ensemble of likely parameters using the JuPOETs multiobjective optimization framework.
The model population was trained and cross-validated to prescribe biological significance using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cell lines \citep{Medici:2008fk}.
The absence of TGF$-\beta$1/2 or VEGF-A stimulation was used as the baseline for the robustness calculations.
Analysis of this population predicted possible phenotypic modes (and their associated signaling) that cells could exhibit when stimulated with TGF-$\beta$ and/or VEGF-A.
The most novel hypothesis generated from the analysis was that cells could operate in a hybrid state defined by both epithelial and mesenchymal traits when stimulated simultaneously with TGF-$\beta$ and VEGF-A.
We tested this hypothesis in MCF10A and DLD1 cells stimulated with combinations of TGF-$\beta$ and VEGF-A.
As expected, in the presence of TGF-$\beta$ or VEGF-A alone, MCF10A and DLD1 cells were either mesenchymal or epithelial, respectively.
However, with both TGF-$\beta$ and VEGF-A, MCF10A and DLD1 cells exhibited a hybrid phenotype, having both epithelial and mesenchymal characteristics.
Furthermore, we found that functional traits such as tubulogenesis and ductal branching were different for cells in this hybrid phenotype.
Together, this study established a predictive model of EMT induction, determined that deterministic model ensembles could predict population heterogeneity, and proved the existence of a unique hybrid phenotype resulting from the simultaneous integration of extracellular growth factor signals.

@@

The deterministic population of EMT models predicted heterogeneous behavior that was qualitatively consistent with experimental studies.
There is a diversity of algorithmic approaches to estimate model parameters \citep{Moles:2003aa}, as well as many strategies to integrate model identification with experimental design \citep{Rodriguez-Fernandez:2013aa,Villaverde:2014aa}.
However, despite these advances, the identification of models describing intracellular network behavior remains challenging.
There are different schools of thought to deal with this challenge.
One school has focused on model reduction.
Data-driven approaches \citep{Cirit:2012kx}, boolean \citep{Choi:2012bs} or other logical model formulations \citep{Morris:2011ys,Terfve:2012kq}
are emerging paradigms that constrain model complexity by the availability of the training and validation data.
Other techniques such as constraints based modeling, which is commonly used to model metabolic networks, have also been applied to model transcriptional networks, although primarily in lower eukaryotes and prokaryotes \citep{Hyduke:2010rt}. These techniques (and many others, see review \citep{Wayman2013}) are certainly exciting, with many interesting properties.
However, we used the traditional approach of mass action kinetics within an ordinary differential equation framework.
The identification problem for the EMT model was massively underdetermined.
This is not uncommon for differential equation models, especially those that are highly mechanistic.
Of course, we could have discarded mechanism or reduced the model scope to decrease the complexity of the identification problem.
However, a central criticism leveled by biologists is that model simplification is often done at the cost of biological reality, or done for reasons of computational expediency \citep{Sainani2012}.
To avoid this criticism, we systematically identified an ensemble of likely models each consistent with the training data, instead of a single but uncertain best fit model.
Previously, we (and others) have suggested that deterministic ensembles could model heterogeneous populations in situations where stochastic computation was not feasible \citep{Lequieu:2011fj}.
Population heterogeneity using deterministic model families has previously been explored for bacterial growth in batch cultures \citep{Lee:2009fk}.
In that case, distributions were generated because the model parameters varied over the ensemble, i.e., extrinsic noise led to population heterogeneity.
In this study, parameters controlling physical interactions such as disassociation rates or the rate of assembly or degradation of macromolecular machinery such as ribosomes were widely distributed over the ensemble. Population heterogeneity can also arise from intrinsic thermal fluctuations, which are not captured by a deterministic population of models \citep{Swain:2002ve}.
Thus, deterministic ensembles, provide a coarse-grained or extrinsic-only ability to simulate population diversity. Despite this limitation, our prediction of phenotypic heterogeneity
(and the underlying signaling events responsible for the heterogeneity) was consistent with experimental observations. This suggested that deterministic ensembles could simulate
disease or developmental processes in which heterogeneity plays an important role, without having to resort to stochastic simulation.

@@

Our signaling network reconstruction was based on Medici et al. who identified the pathways though which MDCKII, DLD1 colon carcinoma,
and A375 melanoma cells transition towards a mesenchymal phenotype \citep{Medici:2008fk}.
Sequential activation of MAPK and Smad pathways were initiated upon addition of TGF$-\beta$1/2.  Briefly, TGF$-\beta$2 signals through the RAS-RAF-MEK-ERK pathway to up-regulate Snail and Slug expression \citep{Medici:2006qa}.  Snail, a known repressor of junctional proteins, inhibits the expression of E-cadherin \citep{Cano:2000kh}. This initial repression of E-cadherin leads to a release of $\beta$-catenin from the cell membrane. This release of $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$-\beta$3 expression \citep{Medici:2008fk}. The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \c\
itep{Medici:2008fk}. GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway
\citep{Zhou:2004kx,Larue:2005vn}. Thereby, further $\beta$-catenin release also resulted from by TGF$-\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \citep{Derynck:2003fc}. This activates the receptors serine/threonine kinase activity to phosphorylate and activate the receptor Smads 2/3 \citep{Massague:2005qc}. In the model, receptors are represented as either bound or unbound complexes.
Phosphorylated Smads 2/3 (pSmad2/3) form heterodimers and translocate to the nucleus. pSmads complexes up-regulate other transcription factors, such as LEF-1. The pSmad2/4-LEF-1 complex has been shown to directly repress the E-cadherin gene \citep{Nawshad:2007kl}. LEF-1 also binds with $\beta$-catenin to upregulate mesenchymal proteins such as fibronectin \citep{Medici2011}.
In the model, Smad signaling is lumped into a single smad species that can act in a co-dependent fashion with LEF1 to downregulate E-cadherin.
The EMT gene expression program was initiated by the binding of TGF$-\beta$ isoforms to TGF$-\beta$ surface receptors.
Binding of extracellular TGF$-\beta$1/2 with TGF$-\beta$ surface receptors I/II (TGF$-\beta$R-I/II) initiates the assembly of adapter complexes which starts the
downstream signaling program.
Complex assembly activates the serine/threonine kinase activity on the receptor, leading to the recruitment and phosphorylation of Smad partners \citep{Massague:2005qc}.
Phosphorylated Smads2/3 (pSmad2/3) form heterodimers with partner Smad4 and then translocate to the nucleus where they act as both transcriptional activators and repressors.
Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \citep{Cano:2000kh}.
However, this transition is not solely driven by transcriptional events.
At the protein level, the repression of E-cadherin leads to a release of $\beta$-catenin from cell membrane.
Cystolic $\beta$-catenin then translocates to the nucleus and forms transcriptionally-active complexes with immunoglobulin transcription factor 2 (TCF-4)
to drive TGF$-\beta$3 expression \citep{Medici:2008fk}. The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \citep{Medici:2008fk}. GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway.
\citep{Zhou:2004kx,Larue:2005vn}. Lastly, VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression \citep{Suehiro:2014aa},
as supported by our VEGF-A experimental data (Fig. \ref{fg:VEGFA-Data}).

@@

\begin{figure}
\includegraphics [width=1.0\linewidth] {./figs/Fig2_training.pdf}
\caption{Training and validation simulations. The population of EMT models qualitatively captured TGF$-\beta$-induced EMT signaling.
(A-I) The population was generated using JuPOETs and trained using 11 different objective functions (41 data sets) taken from Medici \emph{et al.} \citep{Medici:2008fk}.
The model captured the simulated experiments for 72\% of the cases.
(J-L)  The model populations were also compared against untrained temporal data to measure the effectiveness as a pure prediction.}\label{fg:F2}
\end{figure}

@@

\begin{figure}
\includegraphics [width=1.0\linewidth] {./figs/Fig3_phenotype.pdf}
\caption{Simulated VEGF-A and TGF$-\beta$1/2 exposure promoted phenotype heterogeneity.
Simulated response to TGF$-\beta$1/2 and VEGF-A exposure with and without axis specific inhibitors.
Vimentin and E-cadherin abundances (in nM) were used to quantify the shift in population at 48 hrs.
(A-C) VEGF-A (50 a.u.) treatment resulted in a population with enhanced epithelial properties.
This was contrary to the addition of TGF$-\beta$2 (10 a.u.), which shifted the population towards a mesenchymal phenotype.
Interestingly, the combined effects of TGF$-\beta$2 and VEGFA was found to increase both ecadherin and vimentin levels, creating a heterogeneous population (black arrow), which can also be seen in a minority of untreated cells (gray arrow). (D-F) To isolate the effect of NFAT, we inhibited NFAT de-phosphorylation in combination with VEGFA. This negated the increase in ecadherin expression and shifted the population towards a mesenchymal phenotype (Q1,Q3).  Likewise, combining NFAT inhibition with TGF$-\beta$ mitigated all VEGF enhanced ecadherin expression.
Lastly, combination of TGF$-\beta$2, VEGFA, and NFAT inhibition nearly mitigated all effects of VEGFA, shifting the heterogeneous population towards a mesenchymal phenotype.
In whole, high levels of phosphorylated-Sp1 correlated with vimentin expression, while NFAT was responsible for maintaining E-cadherin expression, although neither were mutually exclusive.}
\label{fg:F3}
\end{figure}

>>>>

\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
\vspace{0.05in}
{\par \centering Russell Gould$^{\ddag}$ and David M. Bassen, Anirikh Chakrabarti, Jeffrey D. Varner and Jonathan Butcher$^{\ddag *}$}
\vspace{0.05in}
{\par \centering Robert Frederick Smith School of Chemical and Biomolecular Engineering and $^{\ddag}$Nancy E. and Peter C. Meinig School of Biomedical Engineering, Cornell University, Ithaca NY 14853}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~ Modeling of TGF-$\beta$ induced EMT}
\vspace{0.1in}
% {\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology (60,000 Character Limit)}}
%\vspace{0.1in}
{\par \centering \textbf{Character count:}~\emph{56,659}}
\vspace{0.1in}
{\par \centering $^{*}$Corresponding author:}
\vspace{0.1in}
{\par \centering Jonathan T. Butcher,}
{\par \centering Associate Professor, Nancy E. and Peter C. Meinig School of Biomedical Engineering,}
{\par \centering 304 Weill Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jtb47@cornell.edu}
{\par \centering Phone: (607) 255 - 3575}
{\par \centering Fax: (607) 255 - 7330}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
\section*{Abstract}
Epithelial to mesenchymal transition (EMT) is an essential differentiation program during tissue morphogenesis and remodeling.
EMT is induced by soluble transforming growth factor $\beta$ (TGF-$\beta$) family members, and restricted by vascular endothelial growth factor family members.
While many downstream molecular regulators of EMT have been identified, these have been largely evaluated individually without considering potential crosstalk.
In this study, we created an ensemble of dynamic mathematical models describing TGF-$\beta$ induced EMT to better understand the operational hierarchy of this complex molecular program.
These models incorporate mass action kinetics within an ordinary differential equation (ODE) framework to describe the transcriptional and post-translational regulatory events driving EMT.
Model parameters were estimated from multiple data sets using multiobjective optimization, in combination with cross-validation.
TGF-$\beta$ exposure drove the model population toward a mesenchymal phenotype, while an epithelial phenotype was enhanced following vascular endothelial growth factor A (VEGF-A) exposure.
Simulations predicted that the transcription factors phosphorylated SP1 and NFAT were master regulators promoting or inhibiting EMT, respectively.
Surprisingly, simulations also predicted that a cellular population could exhibit phenotypic heterogeneity (characterized by a significant fraction of the population with both high epithelial and mesenchymal marker expression)
if treated simultaneously with TGF-$\beta$ and VEGF-A.  We tested this prediction experimentally in both MCF10A and DLD1 cells and found that upwards of 45\% of the cellular population acquired this hybrid state in the presence of both TGF-$\beta$ and VEGF-A.
We experimentally validated the predicted NFAT/Sp1 signaling axis for each phenotype response. Lastly, we found that cells in the hybrid state had significantly different functional behavior when compared to VEGF-A or TGF-$\beta$ treatment alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and potentially reveal a novel signaling axis which regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

We identified model subpopulations that exhibited different behaviors following exposure to TGF$-\beta$1/2 (Fig. \ref{fg:F3}B).
Analysis of the molecular signatures of these subpopulations suggested the abundance, localization and state of the Sp1, AP-1 and NFATc transcription factors controlled population heterogeneity.
The majority of models ($>$80\%) responded to treatment, moving away from the untreated population (Fig \ref{fg:F3}A-F, gray).
These models showed the classically expected behavior, a switch from an epithelial to mesenchymal phenotype following TGF$-\beta$1/2 exposure.
Some models resembled untreated cells; they had elevated phosphorylated Sp1, relative to non-induced cells,
which decreased E-cadherin expression through Slug-mediated inhibition, which in turn increased Vimentin expression through TGF$-\beta$3 autocrine signaling and the liberation of $\beta$-catenin. However, the most biologically interesting behavior was exhibited by cells achieving a hybrid phenotype, most notable in a dual treatment condition (\ref{fg:F3}C, black arrow), but also present in a small percentage of untreated cells (Fig. \ref{fg:F3}B, gray arrow).
Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in simultaneously increased Vimentin and E-cadherin expression (Fig. \ref{fg:F4}A).

@@

\section*{Discussion}
In this study, we developed a family of mechanistic models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of VEGF-A.
The signaling architecture encoded in the model, which contained 74 molecular species interconnected by 169 interactions,
described the expression of 23 genes in response to growth factor stimulation.
This simulation incorporates an unprecedented level of detail compared to previous models, but as a consequence created a large number of unknown model parameters.
Because these parameters could not be estimated uniquely apriori, we estimated an ensemble of likely parameters using the JuPOETs multiobjective optimization framework.
The model population was trained and cross-validated to prescribe biological significance using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cell lines \citep{Medici:2008fk}.
Analysis of this population predicted possible phenotypic modes (and their associated signaling) that cells could exhibit when stimulated with TGF-$\beta$ and/or VEGF-A.
The most novel hypothesis generated from the analysis was that cells could operate in a hybrid state defined by both epithelial and mesenchymal traits when stimulated simultaneously with TGF-$\beta$ and VEGF-A.
We tested this hypothesis in MCF10A and DLD1 cells stimulated with combinations of TGF-$\beta$ and VEGF-A.
As expected, in the presence of TGF-$\beta$ or VEGF-A alone, MCF10A and DLD1 cells were either mesenchymal or epithelial, respectively.
However, with both TGF-$\beta$ and VEGF-A, MCF10A and DLD1 cells exhibited a hybrid phenotype, having both epithelial and mesenchymal characteristics.
Furthermore, we found that functional traits such as tubulogenesis and ductal branching were different for cells in this hybrid phenotype.
Together, this study established a predictive model of EMT induction, determined that deterministic model ensembles could predict population heterogeneity, and proved the existence of a unique hybrid phenotype resulting from the simultaneous integration of extracellular growth factor signals.

@@

The deterministic population of EMT models predicted heterogeneous behavior that was qualitatively consistent with experimental studies.
There is a diversity of algorithmic approaches to estimate model parameters \citep{Moles:2003aa}, as well as many strategies to integrate model identification with experimental design \citep{Rodriguez-Fernandez:2013aa,Villaverde:2014aa}.
However, despite these advances, the identification of models describing intracellular network behavior remains challenging.
There are different schools of thought to deal with this challenge.
One school has focused on model reduction.
Data-driven approaches \citep{Cirit:2012kx}, boolean \citep{Choi:2012bs} or other logical model formulations \citep{Morris:2011ys,Terfve:2012kq}
are emerging paradigms that constrain model complexity by the availability of the training and validation data.
Other techniques such as constraints based modeling, which is commonly used to model metabolic networks, have also been applied to model transcriptional networks, although primarily in lower eukaryotes and prokaryotes \citep{Hyduke:2010rt}. These techniques (and many others, see review \citep{Wayman2013}) are certainly exciting, with many interesting properties.
Here, we used a traditional approach of mass action kinetics within an ordinary differential equation framework that also included transfer functions to simplify scenarios where reactions involving one species are controlled by several others (e.g. E-cadherin transcription).
The identification problem for the EMT model was massively underdetermined.
This is not uncommon for differential equation models, especially those that are highly mechanistic.
Of course, we could have discarded mechanism or reduced the model scope to decrease the complexity of the identification problem.
However, a central criticism leveled by biologists is that model simplification is often done at the cost of biological reality, or done for reasons of computational expediency \citep{Sainani2012}.
To avoid this criticism, we systematically identified an ensemble of likely models each consistent with the training data, instead of a single but uncertain best fit model.
Previously, we (and others) have suggested that deterministic ensembles could model heterogeneous populations in situations where stochastic computation was not feasible \citep{Lequieu:2011fj}.
Population heterogeneity using deterministic model families has previously been explored for bacterial growth in batch cultures \citep{Lee:2009fk}.
In that case, distributions were generated because the model parameters varied over the ensemble, i.e., extrinsic noise led to population heterogeneity.
In this study, parameters controlling physical interactions such as disassociation rates or the rate of assembly or degradation of macromolecular machinery such as ribosomes were widely distributed over the ensemble. Population heterogeneity can also arise from intrinsic thermal fluctuations, which are not captured by a deterministic population of models \citep{Swain:2002ve}.
Thus, deterministic ensembles, provide a coarse-grained or extrinsic-only ability to simulate population diversity. Despite this limitation, our prediction of phenotypic heterogeneity
(and the underlying signaling events responsible for the heterogeneity) was consistent with experimental observations. This suggested that deterministic ensembles could simulate
disease or developmental processes in which heterogeneity plays an important role, without having to resort to stochastic simulation.

@@

Our signaling network reconstruction was based on Medici et al. who identified the pathways though which MDCKII, DLD1 colon carcinoma,
and A375 melanoma cells transition towards a mesenchymal phenotype \citep{Medici:2008fk}.
Sequential activation of MAPK and Smad pathways were initiated upon addition of TGF$-\beta$1/2.  Briefly, TGF$-\beta$2 signals through the RAS-RAF-MEK-ERK pathway to up-regulate Snail and Slug expression \citep{Medici:2006qa}.  Snail, a known repressor of junctional proteins, inhibits the expression of E-cadherin \citep{Cano:2000kh}. This initial repression of E-cadherin leads to a release of $\beta$-catenin from the cell membrane. This release of $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$-\beta$3 expression \citep{Medici:2008fk}. The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \citep{Medici:2008fk}. GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway \citep{Zhou:2004kx,Larue:2005vn}. Thereby, further $\beta$-catenin release also resulted from by TGF$-\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \citep{Derynck:2003fc}. This activates the receptors serine/threonine kinase activity to phosphorylate and activate the receptor Smads 2/3 \citep{Massague:2005qc}. In the model, receptors are simplified and represented as either bound or unbound complexes with their ligands.
Phosphorylated Smads 2/3 (pSmad2/3) form heterodimers and translocate to the nucleus. pSmads complexes up-regulate other transcription factors, such as LEF-1. The pSmad2/4-LEF-1 complex has been shown to directly repress the E-cadherin gene \citep{Nawshad:2007kl}. LEF-1 also binds with $\beta$-catenin to upregulate mesenchymal proteins such as fibronectin \citep{Medici2011}.
In the model, Smad signaling is consolidated into a single Smad species that can act in a co-dependent fashion with LEF1 to downregulate E-cadherin via a transfer function, eliminating the need for an explicity LEF-1, pSmad complex.
The EMT gene expression program was initiated by the binding of TGF$-\beta$ isoforms to TGF$-\beta$ surface receptors, starting the downstream signaling program.
Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \citep{Cano:2000kh}.However, this transition is not solely driven by transcriptional events.
At the protein level, the repression of E-cadherin leads to a release of $\beta$-catenin from cell membrane.
Cystolic $\beta$-catenin then translocates to the nucleus and forms transcriptionally-active complexes with immunoglobulin transcription factor 2 (TCF-4)
to drive TGF$-\beta$3 expression \citep{Medici:2008fk}. The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \citep{Medici:2008fk}. GSK3 is known to act on $\beta$-catenin signaling through APC complex associated ubiquitin-proteasome pathway. The APC complex is represented in our model and serves as a second reservoir  of $\beta$-cateinin in untransformed cells whose sequestration is regulated by GSK3 \citep{Zhou:2004kx,Larue:2005vn,Medici:2008fk}. Lastly, VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression \citep{Suehiro:2014aa},
as supported by our VEGF-A experimental data (Fig. \ref{fg:VEGFA-Data}).

@@

\begin{figure}
\includegraphics [width=1.0\linewidth] {./figs/Fig2_training.pdf}
\caption{Training and validation simulations. The population of EMT models qualitatively captured TGF$-\beta$-induced EMT signaling.
(A-I) The population was generated using JuPOETs and trained using 11 different objective functions (41 data sets) taken from Medici \emph{et al.} \citep{Medici:2008fk}.
The model captured the simulated experiments for all cases when compared to randomized controls.
(J-L)  The model populations were also compared against untrained temporal data to measure the effectiveness as a pure prediction.}\label{fg:F2}
\end{figure}

@@

\begin{figure}
\includegraphics [width=1.0\linewidth] {./figs/Fig3_phenotype.pdf}
\caption{Simulated VEGF-A and TGF$-\beta$1/2 exposure promoted phenotype heterogeneity.
Simulated response to TGF$-\beta$1/2 and VEGF-A exposure with and without axis specific inhibitors.
Vimentin and E-cadherin abundances (in nM) were used to quantify the shift in population at 48 hrs.
(A-C) VEGF-A (50 a.u.) treatment resulted in a population with enhanced epithelial properties.
This was contrary to the addition of TGF$-\beta$2 (10 a.u.), which shifted the population towards a mesenchymal phenotype.
Interestingly, the combined effects of TGF$-\beta$2 and VEGFA was found to increase both ecadherin and vimentin levels, creating a heterogeneous population (black arrow), which can also be seen in a minority of untreated cells (gray arrow). (D-F) To isolate the effect of NFAT, we inhibited NFAT de-phosphorylation in combination with VEGFA. This negated the increase in ecadherin expression and shifted the population towards a mesenchymal phenotype.  Likewise, combining NFAT inhibition with TGF$-\beta$ mitigated all VEGF enhanced ecadherin expression.
Lastly, combination of TGF$-\beta$2, VEGFA, and NFAT inhibition nearly mitigated all effects of VEGFA, shifting the heterogeneous population towards a mesenchymal phenotype.
In whole, high levels of phosphorylated-Sp1 correlated with vimentin expression, while NFAT was responsible for maintaining E-cadherin expression in the presence of other factors, although neither were mutually exclusive.}
\label{fg:F3}
\end{figure}
